---
title: "Oxygen diffusion and respiratory consumption in lung mucus"
output: html_document
---

All source code and data needed to generate this report is available [online](https://github.com/sebkopf/o2cf).

```{r load-funcs, message=FALSE, warning=FALSE}
source("functions.R")
source("model.R")
```

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
# load all dependencies
packages <- c(
  "knitr", # knitr package for rendering the markdown file
  "ggplot2", # plotting 
  "scales", # for scale notations
  "RColorBrewer", # colour palettes for plotting
  "grid", # plot structuring
  "reshape2", # data frame operations
  "plyr", # data frame operations
  "deSolve", # differential equation solver
  "ReacTran", # diffusion modeling
  "rootSolve", # steady state calculations
  "lubridate", # for converting times to most accessible time units
  "xlsx", # read excel files (requires Java runtime)
  "testthat" # for testing code
)
load_pkgs(packages)

# knitr options
opts_knit$set(root.dir=".")
opts_knit$set(cache = TRUE)
opts_knit$set(autodep = TRUE)
options(pdfsave.prefix = "plot-")
options(pdfsave.folder = "figures")
```

### Theory

Details on the system are in the main text. Here just a quick summary of the non-dimensionalized differential equations and boundary conditions.

Differential equations:

 - along a line (column): $\frac{\partial{u}}{\partial{t}} = \left(  \frac{\partial{}^2 u}{\partial{x^2}} \right) - \left( \frac{u}{1 + u} + g \right) = 0$
 - radial (ring): $\frac{\partial{u}}{\partial{t}} = \left(  \frac{\partial{}^2 u}{\partial{r^2}} + \frac{1}{r}\frac{\partial{u}}{\partial{r}} \right) - \left( \frac{u}{1 + u} + g \right) = 0$ (assuming radial symmetry, i.e. $\frac{\partial{u}}{\partial{\theta}} = 0$)
 - spherical: $\frac{\partial{u}}{\partial{t}} = \left(  \frac{\partial{}^2 u}{\partial{r^2}} + \frac{2}{r}\frac{\partial{u}}{\partial{r}} \right) - \left( \frac{u}{1 + u} + g \right) = 0$ (assuming radial and azimuthal symmetry, i.e. $\frac{\partial{u}}{\partial\theta}=0\,;\,\frac{\partial{u}}{\partial\phi}=0$)

Boundary conditions (difference between outward vs. inward only applicable to radial and spherical geometry):

 - outward diffusion:
    - $u(r = r_1) = {O_2}_{max}$ (oxygen maximal at the inner air-mucus boundary, diffusion across the air-water interface much faster than diffusion deeper into the mucus so this is treated as constant supply)
    - $\left.\frac{\partial{u}}{\partial{r}} \right|_{r=r_2} = 0$  (reflecting boundary at the outer boundary, for example the mucus-ephilial cell surface)
 - inward diffusion:
    - $u(r = r_2) = {O_2}_{max}$ (oxygen maximal at the outer boundary)
    - $\left.\frac{\partial{u}}{\partial{r}} \right|_{r=r_1} = 0$  (inner boundary reflecting)

Diffusion is modeled using the **ReacTran** package (Soetaert and Meysman, 2012) and steady state concentrations ($\frac{\partial{u}}{\partial{t}} == 0$) are computed using the **rootSolve** package (Soetaert, 2009).

### Prep

Make sure default parameters are loaded and the non-dimensionalization and modeling functions are running as expected.

#### Default parameters

Load default parameters from the constants excel file.

```{r defaults, results='asis'}
constants <- read.xlsx("parameters.xlsx", sheetName = "constants", stringsAsFactors = F, header=TRUE, check.names=FALSE)
kable(constants)
```

#### Testing

Running a few tests to make sure code and parameter changes don't break the code in unexpected ways. None of these tests should throw an error if the non-dimensionalizatiis working as expected.

```{r tests}
test_that("forward backward non-dimensionalizatoin / re-dimensionalization works", {
  expect_equal(re_dim_o2(non_dim_o2(200)), 200)
  expect_equal(re_dim_x(non_dim_x(100)), 100)
  expect_equal(re_dim_t(non_dim_t(20)), 20)
})

test_that("scaling and non-dimensional evaluation matches independently computed values", {
  expect_equal(signif(calc_spatial_scaling_factor(), 5), 39860)
  expect_equal(signif(calc_temporal_scaling_factor(), 5), 2.4271)
  expect_equal(signif(calc_maintenance_term(), 5), 0.010921)
  expect_equal(signif(non_dim_o2(250), 5), 20.161)
  expect_equal(signif(non_dim_x(100), 5), 3.986)
  expect_equal(signif(non_dim_t(60), 5), 145.63)
})
```

## Modeling

The O2 diffusion vs. respiration modeling. Load all modeling parameters:

```{r load-model-params, results='asis'}
params <- read.xlsx("parameters.xlsx", sheetName = "models", stringsAsFactors = F, header=TRUE, check.names=TRUE)
kable(params)
```


### Oxygen penetration into mucus in different scenarios

Compare oxygen penetration into a mucus clogged airway, disloged mucus aggregate (presumed spherical) and the mucus lining of partially clogged airway

#### Run model

```{r run-model1, results='asis'}
data1 <- ddply(subset(params, test == "geometry"), names(params), function(df) {
  with(unique(df), {
    result <- model_o2(
      x_min = free.um, x_max = free.um + mucus.um, geometry = geometry, 
      direction = direction, cells = cell.pow, D_o2 = D_o2)
    if (direction == "inward")
      return(mutate(result, x = max(x) - x))
    else
      return(mutate(result, x = x - min(x)))
    })
})

kable(summarize_model_results(data1, params))
```

#### Plot geometry comparison

```{r oxygen_profiles_different_geometries, fig.width=10, fig.height=7, save_pdf=TRUE}
p1 <- ggplot(data1) +
  geom_rect(data = data.frame(layer = c("air", "mucus"), min = c(-50, 0), max = c(0, Inf)), 
            aes(xmin = -Inf, xmax = Inf, ymin = min, ymax = max, fill = layer), alpha = 0.1) +
  geom_line(aes(o2, x, colour = factor(cell.pow)), size = 1) + 
  geom_hline(yintercept = 0, colour = "black", size = 1, linetype = 2) + 
  scale_colour_manual ("Microbial load: ", values = brewer.pal(9, "Set1"), 
                       labels = function(value) sapply(value, function(i) bquote("10"^.(i)~"cells/ml"))) + 
  scale_fill_manual("Layer: ", values = c("#2171B5", "#FEE08B")) + 
  scale_x_continuous(expression("O"[2]~"["*mu*M*"]"), expand=c(0,0)) +
  scale_y_reverse(expression("Depth into mucus ["*mu*m*"]"), expand = c(0, 0)) + 
  facet_grid(~Scenario, labeller = function(var, val) { 
      sapply(as.character(val), function(i) {
          lines <- strsplit(i, "\\\\\\\\")[[1]]
          two_line_label(lines[1], lines[2])
        })
    }) + 
  theme_bw() + theme_xangle() + theme_facet() + theme_public() + 
  theme(legend.position = "bottom") 
p1
```

### Comparision between differen diffusion coefficients

Here we compare the effects of variable diffusion coefficients.

#### Run model

```{r run-model2, results='asis'}
data2 <- ddply(subset(params, test == "diffusivity"), names(params), function(df) {
  with(unique(df), model_o2(x_min = free.um, x_max = free.um + mucus.um, 
                            geometry = geometry, direction = direction, 
                            cells = cell.pow, D_o2 = D_o2))
})

kable(summarize_model_results(data2, params))
```

#### Plot diffusion comparison

```{r oxygen_profiles_variable_diffusivities, fig.width=10, fig.height=7, save_pdf=TRUE}
p1 %+% data2  %+%
  facet_grid(~D_o2, labeller = function(variable, value){
    sapply(value, function(i) 
      two_line_label(unique(params$Diffusion[params$D_o2 == i]), 
                     bquote("D"["O2"]*":"~.(round(i, 1))%.%10^-6~cm^2~s^-1)))
  }) 
```


### Oxycline depth

Find the minimal mucus thickness at which oxygen concentrations reach below the cutoff of `r get_default("o2_cutoff", "µM")` µM (i.e. find the thickness of the oxycline).

```{r o2-depth, results='asis'}
cutoff_depths <- ddply(subset(params, test == "cutoff"), names(params), function(df) {
  with(unique(df), find_mucus_thickness(
    o2 = get_default("o2_cutoff", "µM"), 
    start_x = mucus.um, # use mucus.um as first guess
    geometry = geometry, direction = direction, 
    cells = cell.pow, D_o2 = D_o2))
})

# summarize
cutoff_info <- dcast(
  mutate(cutoff_depths,
         `Density [cells/ml]` = scientific(10^cell.pow, digits=2),
         rounded = signif(depth, 2),
         value = ifelse(rounded > 1000, paste0(rounded/1000, " mm"), paste0(rounded, " µm"))
         ),  cell.pow + `Density [cells/ml]` ~ Scenario, value.var = "value")[-1]

if (!file.exists("tables")) dir.create("tables")
write.csv(cutoff_info, file = "tables/table_oxycline_model.csv", row.names = FALSE)
kable(cutoff_info)
```


### Diffusion into mucus lining

Here we visualize how the thickness of the mucus layer inside an airway can change the oxygen profiles.


#### Run model

```{r run-model3, results='asis'}
data3 <- ddply(subset(params, test == "thickness"), names(params), function(df) {
  with(unique(df), model_o2(x_min = free.um, x_max = free.um + mucus.um, 
                            geometry = geometry, direction = direction, 
                            cells = cell.pow, D_o2 = D_o2))
})

kable(summarize_model_results(data3, params))
```


##### Plot profile

Plot a single transect profile from the center of a bronchiole to the mucus-epiphelial cell boundary.

```{r oxygen_profiles_bronchiole_transect, fig.width=8, fig.height=5.5, save_pdf=TRUE}
# facet grid labeler
grid_labeller <- function(variable,value){
  if (variable == "mucus.um")
    return (sapply(value, function(i) bquote("mucus:"~.(i)~mu*"m")))
  else if (variable == "cell.pow")
    return (sapply(value, function(i) bquote("10"^.(i)~"cells/ml")))
}

# plot transect
ggplot(data3) +
  geom_rect(data = unique(mutate(subset(params, test == "thickness", sel=c("free.um", "mucus.um")), layer = "mucus-free airway")), 
            aes(min = -Inf, ymax = Inf, xmin = 0, xmax = free.um, fill = layer), alpha = 0.1) +
  geom_rect(data = unique(mutate(subset(params, test == "thickness", sel=c("free.um", "mucus.um")), layer = "mucus layer")), 
            aes(min = -Inf, ymax = Inf, xmin = free.um, xmax = free.um + mucus.um, fill = layer), alpha = 0.1) +
  geom_line(aes(x, o2, colour = factor(cell.pow)), size = 1) + 
  geom_hline(yintercept = get_default("o2_max", "µM"), colour = "black", size = 1, linetype = 2) + 
  scale_colour_manual ("Microbial load: ", values = brewer.pal(9, "Set1"), 
                       labels = function(value) sapply(value, function(i) bquote("10"^.(i)~"cells/ml"))) + 
  scale_fill_manual("Layer: ", values = c("#FEE08B", "#2171B5")) + 
  scale_x_continuous(expression("Distance from center of bronchiole ["*mu*m*"]"), expand=c(0,0)) +
  facet_grid(~mucus.um, labeller = grid_labeller) + 
  labs(y = expression("O"[2]~"["*mu*M*"]")) +
  theme_bw() + theme_xangle() + theme_facet() + theme_public() + 
  theme(legend.position = "bottom") 
```

### Plot 2D visualization

Visualize the entire bronchiole cross section.

```{r oxygen_profiles_bronchiole_cross_section, fig.width = 10, fig.height = 7, save_pdf = TRUE}
# convert to polar ccoordinates
data_cart <- ddply(data3, .(mucus.um, cell.pow), 
                   function(df) convert_polar_to_cartesian(df$x, df$o2, N=400))

# make o2 plot
spectral_colors <- colorRampPalette (brewer.pal (11, "Spectral")) 
ggplot(data_cart, 
       aes (x, y, fill = z, z = z)) + 
  geom_raster (hjust = 0, vjust = 0) + theme_bw () + 
  scale_fill_gradientn (expression("O"[2]~"["*mu*M*"]"), colours = spectral_colors (4), limits=c(0, get_default("o2_max", "µM"))) + 
  scale_x_continuous(expression("Distance from center of bronchiole ["*mu*m*"]"), labels = function(x) abs(x)) +
  scale_y_continuous(expression("Distance from center of bronchiole ["*mu*m*"]"), labels = function(x) abs(x)) +
  facet_grid(mucus.um~cell.pow, labeller = grid_labeller) + 
  theme_bw() + theme_nogrid() + theme_facet(0) + theme_public() + theme_xangle(angle = 90) + 
  guides(fill = guide_colourbar(barwidth = 3, barheight = 10, label.hjust = 0))
```


### Growth rates

Calculate and visualize growth rates for oxygen limited organisms (assuming monod-type growth rates as a function of oxygen concentrations). Only applies to O2-respiring organisms in the mucus, not to fermenters or anaerobically growing microbes.

```{r growth_rates_bronchiole_cross_section, fig.width = 10, fig.height = 7, save_pdf=TRUE}
# calculate growth rates (in 1/hr)
data_growth <- mutate(data_cart, mu.hr = calculate_monod_growth_rate (z))

# define meaningful breaks in the growth rates and root-scale colour scale to spread the 
# gradient across the breaks (rather than having all colour transitions at one extreme)
dblts.breaks <- list(c(45, "minutes"), c(90, "minutes"), c(5, "hours"), c(1, "day"), c(14, "days"))
dblts.hr.breaks = sapply(dblts.breaks, function(x) duration(as.integer(x[1]), x[2])) / (60 * 60)
mu.hr.breaks <- c(0, log(2)/dblts.hr.breaks)
break.labels <- c(two_line_label("no growth by", O[2]~"respiration"), duration_label(dblts.hr.breaks, "hours"))
colour_scaling <- function(x) x^(1/4) 

# make growth plot
ggplot(data_growth, aes (x, y, fill = colour_scaling(mu.hr))) + 
  geom_raster (hjust = 0, vjust = 0) + theme_bw () + 
  scale_fill_gradientn (
    two_line_label("Generation time", "(resp. growth)"), colours = spectral_colors (4), 
    breaks = colour_scaling(mu.hr.breaks), limits=c(0, max(colour_scaling(mu.hr.breaks))), labels = break.labels) + 
  scale_x_continuous(expression("Distance from center of bronchiole ["*mu*m*"]"), labels = function(x) abs(x)) +
  scale_y_continuous(expression("Distance from center of bronchiole ["*mu*m*"]"), labels = function(x) abs(x)) +
  facet_grid(mucus.um~cell.pow, labeller = grid_labeller) + 
  theme_bw() + theme_nogrid() + theme_facet(0) + theme_public() + theme_xangle(angle = 90) + 
  guides(fill = guide_colourbar(barwidth = 3, barheight = 10, label.hjust = 0))
```

