---
title: "Oxycline implications for microbial density"
output: html_document
---

All source code and data needed to generate this report is available [online](https://github.com/sebkopf/o2cf).

```{r load-funcs, message=FALSE, warning=FALSE}
source("functions.R")
source("model.R")
source("data.R")
```

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
# load all dependencies
packages <- c(
  "knitr", # knitr package for rendering the markdown file
  "ggplot2", # plotting 
  "scales", # for scale notations
  "RColorBrewer", # colour palettes for plotting
  "grid", # plot structuring
  "reshape2", # data frame operations
  "plyr", # data frame operations
  "deSolve", # differential equation solver
  "ReacTran", # diffusion modeling
  "rootSolve", # steady state calculations
  "lubridate", # for converting times to most accessible time units
  "xlsx", # read excel files (requires Java runtime)
  "testthat" # for testing code
)
load_pkgs(packages)

# knitr options
opts_knit$set(root.dir=".")
opts_knit$set(cache = TRUE)
opts_knit$set(autodep = TRUE)
options(pdfsave.prefix = "plot-")
options(pdfsave.folder = "figures")
```

### Load clinical data

Read in clinical data profiles.

```{r read-data, results='asis'}
data <- 
  rbind(
    read_oxycline_data("profiles.xlsx", sheet = "2013"),
    read_oxycline_data("profiles.xlsx", sheet = "2014")
    )
message("Loaded ", nrow(data), " data points from ", length(unique(data$profile)), " oxygen profiles.")
```

#### Default parameters

Load default parameters from the constants excel file.

```{r params, results='asis'}
constants <- read.xlsx("parameters.xlsx", sheetName = "constants", stringsAsFactors = F, header=TRUE, check.names=FALSE)
kable(constants)
```

Finds start of oxyclines by determining the biggest continuous drop in O2 concentration (i.e. largest $\Delta{}C$ over range where first spatial derivative is negative, $\frac{dC}{dx} < 0$) and finding the local minimum of the 2nd spatial derivative $\left(-\frac{d^2 C}{dx^2}\right)$ in this interval (i.e. the highest gradient change). The oxygen profile is smoothed with a gaussian kernel for this purpose to reduce noise. Finds end of oxyclines by determing when oxygen concentrations first reach below the cutoff of `r get_default("o2_cutoff", "µM")` µM.

```{r find`-profiles}
profiles <- find_midpoints(data, low_o2_threshold = 1)
profiles <- find_oxyclines(profiles, kernel_width = 10, no_o2_cutoff = get_default("o2_cutoff", "µM"))
```

#### Show profiles

Show oxygen profiles with the oxyclines highlighted in red and framed by dashed lines.

```{r data_profiles, results='asis', fig.width = 16, fig.height = 16, save_pdf = TRUE }
plot.df <- mutate(profiles, panel = paste0(format(date, format = "%m/%d/%y"), ", ", patient))
p1 <- ggplot(NULL) + 
  geom_hline(data = unique(plot.df[c("panel", "oxycline_start")]), aes(yintercept = oxycline_start), linetype = 2) +
  geom_hline(data = unique(plot.df[c("panel", "oxycline_end")]), aes(yintercept = oxycline_end), linetype = 2) +
  geom_point(data = subset(plot.df, !is_oxycline), 
             aes(x = conc, y = depth), shape = 21, colour = "black", fill = "white") +
  geom_point(data = subset(plot.df, is_oxycline), 
             aes(x = conc, y = depth, fill = orp), shape = 21, colour = "black") +
  scale_y_reverse(labels = function(x) x/1000) + 
  scale_fill_manual("Redox conditions", values = c("#999999", "#E41A1C", "#377EB8")) +
  theme_bw() + theme_xangle() + theme_facet(margin.cm=0.1) + theme_public() + theme_nogrid() + 
  theme(legend.position = "bottom", legend.direction = "vertical") + 
  labs(y = "Probe depth [mm]", x = expression("O"[2]~"["*mu*M*"]")) +
  facet_wrap(~panel, scales = "free") 
p1
```


### Fit to model

Find best fit cell density for clinical data (from start of oxycline to profile midpoint). Sputum plugs are modeled as idealized spheres with oxygen diffusing in from the surface to the center. 

```{r run, results='asis'}
models <- ddply(profiles, .(profile, date, patient), function(df) {
  find_cell_density_from_oxycline(
    subset(df, (depth >= oxycline_start) & (depth <= midpoint_depth)),
    geometry = "sphere", direction = "inward", D_o2 = get_default("D_o2_biofilm", "10-6 cm2 s-1"))
})
```

Compile an overview for the data including the total profile, oxycline thickness and the modeling results. 

```{r data-overview, results='asis'}
profile_info <- ddply(profiles, .(date, patient), summarize,
      `Total profile [mm]` = (max(depth) - min(depth))/1000,
      `Oxycline thickness [mm]` = (unique(oxycline_end) - unique(oxycline_start))/1000)

model_info <- 
  ddply(models, .(date, patient), summarize, 
      `Time to find best fit` = paste0(signif(unique(elapsed), 3), " sec"),
      `Cell density [cells/ml]` = scientific(10^unique(cell_density), digits=2), 
      `Root-mean-square error [microM]` = paste0(signif(unique(rms), 3)))
data_table <- merge(profile_info, model_info, by = c("date", "patient"))

if (!file.exists("tables")) dir.create("tables")
write.csv(data_table, file = "tables/table_oxycline_fits.csv", row.names = FALSE)
kable(data_table)
```

Cell density calculations summary:

```{r}
message("Cell density from fit [cells/ml]", 
        ": mean=", scientific(mean(10^models$cell_density), d=2),
        "; median=", scientific(median(10^models$cell_density), d=2),
        "; min=", scientific(min(10^models$cell_density), d=2),
        "; max=", scientific(max(10^models$cell_density), d=2))
```


### Visualize results

Plotting model fits in green on top of data

```{r data_model_fits, fig.width = 16, fig.height = 16, save_pdf = TRUE}
p1 + 
  geom_line(
    data = mutate(models, panel = paste0(format(date, format = "%m/%d/%y"), ", ", patient)),
    aes(x = o2, y = model_depth), colour = "#4DAF4A", size = 1, linetype = 1) 
```

